---
title: "BUAN 650 - Final Project"
output:
  html_document:
    df_print: paged
date: "December 2025"
---
# **TABLE OF CONTENTS**

- [Overview](#overview)
- [Data Preparation](#data-preparation)
- [Exploratory Data Analysis](#exploratory-data-analysis)
- [Predictive Modeling](#predictive-modeling)
- [Recommended Solutions](#recommended-solutions)

# **OVERVIEW**

## Introduction
The provided data represents approximately six months of e-commerce activity for a multi-category retailer. The purpose of this project is to: 

1. Understand the data, focusing on customer or product features associated with purchases

2. Determine the impact of a specific problem occurring in a month of data

3. Use models to find patterns that can help marketing or product departments recover lost revenue in future months

## Specific Problem/Challenge
A 10% decline occurs in the cart-to-purchase ratio for a period/month. This implies that there is some issue affecting a user's likelihood to make a purchase after adding an item to their cart.

## Assumptions 
- We use the latest month, February, as the baseline data
- We assume that the decline in purchases is distributed randomly and attempt to understand patterns associated with the decline

## Analysis Process

### Data Preparation Overview
- In order to prepare the data for analysis, we deduplicated and performed some modifications to the data to make it easier to analyze, such as designating user identification as categorical and differentiating product sub-categories, as well as creating multiple versions of the price field in case it was needed in modeling.

### Exploratory Data Analysis Overview
- Using the entire dataset, we sought to understand the current cart-to-purchase ratio and total revenue associated with cart adds and purchases, and we created some descriptive statistics and visualizations. The overall cart-to-purchase ratio appears to be .59, with $8.6M in revenue added to carts and $5.1M in revenue in purchases for the entire dataset.  
- We then extracted a baseline month (February) and simulated March data based on February data with a 10% decline in the cart-to-purchase ratio by randomly removing sufficient purchases to result in that decline.
- In the March data, we examined the impact on lost revenue in terms of several factors: products, pricing, and user engagement features. We then merged fields such as interaction counts, session duration, user interaction frequency, and category diversity back into the dataset for use in modeling.

**Exploratory Data Analysis Summary**

- The cart-to-purchase ratio declined by 10% in March compared to February
- Total revenue for February (baseline): $1,376,806
- Financial impact of the decline: $213,921.06 in lost revenue
- Estimated total revenue for March: $1,162,885
- Top categories affected: Videocards
- The top selling products remained the same videocards from February to March
- Most impacted price buckets: $200–$300

### Predictive Modeling Overview
- The goal of our predictive modeling approach was to understand the features and predict the decision likelihood for cart-to-purchase conversion. With a better sense of why some customers were more likely than others to convert to a purchase, we could then identify the key influencers that product and marketing teams can use to influence conversion behavior and increase revenue. 
- We used classification tree, random forest, and logistic models to classify the likelihood that a session would convert to a purchase and compared the performance of all three models. We chose these models due to their usefulness in classification analysis, as well as their practicality; implementing these models is fairly straightforward, and if the drop in cart-to-purchase ratio is a continuing issue, we wanted to provide relatively easy-to-implement models for the company to use in addressing this issue on an ongoing basis. Other models, like XGBoost, are more resource-intensive.
- Based on several performance metrics, the random forest model was the most successful at identifying the key factors driving cart-to-purchase conversions. This was the model we selected, allowing us to recommend targeted actions to influence conversion, grounded in statistical insights. 
- Then, we simulated a third month in which the higher likelihood sessions are promoted, sufficient to ensure that the total value of purchases increased from $1,162,885 in March to $1,376,806 or more in the next month, recovering lost revenue.
- Additionally, we performed regression analysis to determine the specific revenue impact of increasing an example variable (such as cart additions); however, findings from this model were not chosen for our recommendations given the model's limited robustness.

## Recommendations Overview
Among customers with the highest likelihood of converting from cart to purchase, the key influencers are: 

- **Cart Count:** The number of items added to the cart
- **Session Duration:** The time spent on the platform during a session
- **Views Count:** The number of product views per session
- **Price:** Low- to mid-range items (<$300 USD) are more likely to convert
- **Category Diversity:** Users interacting with multiple categories of products are more likely to convert
- **Product Category:** Specific high-level product categories influence the likelihood of conversion

More detail on specific interventions the company could implement are available in the recommendations section of this report. Additionally, limitations in our analysis are discussed in that section in detail and include assumptions in the distribution of data, incomplete data, model simplifications, and simulation limitations.

In summary, the focus should be on recovering $213,921 in lost revenue through enhanced user engagement, optimized pricing, and category diversity. Immediate actions include cart abandonment campaigns, showcasing bestsellers, and cross-selling prompts, while medium-term strategies involve dynamic pricing, curated promotions, and category exploration. 

A campaign that targets these key influencers and that results in a ~20% increase in top likelihood customers and ~20% decrease in lowest likelihood customers would result in a revenue increase of ~$217,000 from March to April, recovering the lost revenue.

# **DATA PREPARATION**

# **Step 1: Get required libraries and data**

## Step 1a: Get required libraries
```{r}
options(warn = -1)
suppressMessages(library(readxl))
suppressMessages(library(janitor))
suppressMessages(library(rsample))
suppressMessages(library(ROCR))
suppressMessages(library(glmnet))
suppressMessages(library(plotmo))
suppressMessages(library(car))
suppressMessages(library(dplyr))
suppressMessages(library(scales))
suppressMessages(library(tidyr))
suppressMessages(library(rpart))
suppressMessages(library(rpart.plot))
suppressMessages(library(caret))
suppressMessages(library(randomForest))
suppressMessages(library(rpart))
suppressMessages(library(ROCR))
suppressMessages(library(ggplot2))
suppressMessages(library(ggpmisc))
```

## Steb 1b: Get data
```{r}
suppressMessages(setwd("C:/Users/henn5285/OneDrive - University of St. Thomas/Documents/MSBA/Predictive Analytics/Data Files/Final Project/"))

suppressMessages(FinalProjectData <- read_excel("FinalProjectData.xlsx"))
```

# **Step 2: Inspect and clean the data**

## Step 2a: Preliminary step: Create a dataframe
```{r}
df <- FinalProjectData
```

## Step 2b: View data summaries
```{r}
head(df)
summary(df)
```

## Step 2c: Overall data cleaning and preparation

### Data de-duplication
We needed to de-duplicate data where there are rows with all information the same; these are duplicate data entries, not independent events
```{r}
# Identify duplicate rows in the data frame
duplicates <- df[duplicated(df) | duplicated(df, fromLast = TRUE), ]

# View the first few rows of duplicates
head(duplicates)

# Get a summary of duplicates
summary(duplicates)
count(duplicates)

# Remove duplicate rows
deduplicateddf <- distinct(df)
```

### Designate specific columns as character data type
```{r}
deduplicateddf$product_id <- as.character(deduplicateddf$product_id)
deduplicateddf$category_id <- as.character(deduplicateddf$category_id)
deduplicateddf$user_id <- as.character(deduplicateddf$user_id)
```

### Convert UTC timestamp into separate date and time objects
```{r}
timestamp_posix <- as.POSIXlt(deduplicateddf$event_time, tz = "UTC")
deduplicateddf$time_month <- timestamp_posix$mon + 1  # Adding 1 because month index starts from 0
deduplicateddf$time_day <- timestamp_posix$mday
deduplicateddf$time_hour <- timestamp_posix$hour
```

### Split the category column 
```{r}
split_categories <- strsplit(deduplicateddf$category_code, "\\.")
deduplicateddf$category_1 <- sapply(split_categories, function(x) x[1])
deduplicateddf$category_2 <- sapply(split_categories, function(x) ifelse(length(x) > 1, x[2], NA))
deduplicateddf$category_3 <- sapply(split_categories, function(x) ifelse(length(x) > 2, x[3], NA))
```

### Create multiple versions of price field if needed in modeling
```{r}
hist(deduplicateddf$price)
deduplicateddf$price_log = log(deduplicateddf$price)
deduplicateddf$price_sqrt = sqrt(deduplicateddf$price)
hist(deduplicateddf$price_log)
hist(deduplicateddf$price_sqrt)
```

# **Step 3: Exploratory analysis on original, deduplicated data**

## Step 3a: Descriptive Statistics
```{r}
# Summary statistics
summary_stats <- deduplicateddf %>%
  summarise(
    min_price = min(price, na.rm = TRUE),
    max_price = max(price, na.rm = TRUE),
    mean_price = mean(price, na.rm = TRUE),
    median_price = median(price, na.rm = TRUE),
    total_revenue = sum(price, na.rm = TRUE)
  )

print(summary_stats)

# Unique value counts
unique_counts <- deduplicateddf %>%
  summarise(
    unique_products = n_distinct(product_id),
    unique_categories = n_distinct(category_1),
    unique_users = n_distinct(user_id),
    unique_sessions = n_distinct(user_session)
  )

print(unique_counts)
```

## Step 3b: Visualizations of Price, Event Types, Revenue
```{r}
# Distribution of Prices
ggplot(deduplicateddf, aes(x = price)) +
  geom_histogram(binwidth = 10, fill = "blue", color = "black") +
  labs(title = "Distribution of Prices", x = "Price", y = "Frequency") +
  theme_minimal()

# Distribution of Event Types
event_counts <- deduplicateddf %>%
  group_by(event_type) %>%
  summarise(count = n())

ggplot(event_counts, aes(x = event_type, y = count, fill = event_type)) +
  geom_bar(stat = "identity") +
  labs(title = "Event Type Distribution", x = "Event Type", y = "Count") +
  theme_minimal()

# Total Revenue by Event Type
event_revenue <- deduplicateddf %>%
  group_by(event_type) %>%
  summarise(total_revenue = sum(price, na.rm = TRUE))

ggplot(event_revenue, aes(x = event_type, y = total_revenue, fill = event_type)) +
  geom_bar(stat = "identity") +
  labs(title = "Revenue by Event Type", x = "Event Type", y = "Total Revenue") +
  theme_minimal()
```

## Step 3c: Identify mean and cumulative value by event/action type: View, Cart, Purchase
```{r}
actionSums <- tapply(deduplicateddf$price, deduplicateddf$event_type, sum)
actionSums_dollar <- dollar(actionSums)
actionSums_dollar

original_cart_to_purchase_ratio <- 5125114/8625870
original_cart_to_purchase_ratio

actionMean <- tapply(deduplicateddf$price, deduplicateddf$event_type, mean)
actionMeans_dollar <- dollar(actionMean)
actionMeans_dollar
```

## Step 3d: Summarize by user

### Price and price by event types by user
```{r}
summary_table_userpriceoriginal <- deduplicateddf %>%
  group_by(user_id) %>%
  summarise(
    total_actions = n(),                   # Count of actions per user
    total_price = sum(price, na.rm = TRUE), # Total price per user
    price_view = sum(price[event_type == "view"], na.rm = TRUE), # Sum for event type 1
    price_cart = sum(price[event_type == "cart"], na.rm = TRUE), # Sum for event type 2
    price_purchase = sum(price[event_type == "purchase"], na.rm = TRUE)  # Sum for event type 3
  )

summary_table_userpriceoriginal
```

### Cart-to-purchase ratio by user
```{r}
# Calculate cart-to-purchase ratios by user
cart_purchase_ratio_user <- deduplicateddf %>%
  filter(event_type %in% c("cart", "purchase")) %>%
  group_by(user_id, event_type) %>%
  summarise(
    event_count = n(),
    total_price = sum(price, na.rm = TRUE)
  ) %>%
  pivot_wider(names_from = event_type, values_from = c(event_count, total_price), values_fill = 0) %>%
  mutate(
    cart_to_purchase_ratio_count = ifelse(event_count_cart > 0, event_count_purchase / event_count_cart, NA),
    cart_to_purchase_ratio_price = ifelse(total_price_cart > 0, total_price_purchase / total_price_cart, NA)
  )

# View the results
cart_purchase_ratio_user
```

## Step 3e: Summarize by session

### Price and price by event types by session
```{r}
summary_table_sessionpriceoriginal <- deduplicateddf %>%
  group_by(user_session) %>%
  summarise(
    total_actions = n(),                   # Count of actions per user
    total_price = sum(price, na.rm = TRUE), # Total price per user
    price_view = sum(price[event_type == "view"], na.rm = TRUE), # Sum for event type 1
    price_cart = sum(price[event_type == "cart"], na.rm = TRUE), # Sum for event type 2
    price_purchase = sum(price[event_type == "purchase"], na.rm = TRUE)  # Sum for event type 3
  )

summary_table_sessionpriceoriginal
```

### Cart-to-purchase ratio by session
```{r}
# Calculate cart-to-purchase ratios by session
cart_purchase_ratio_session <- deduplicateddf %>%
  filter(event_type %in% c("cart", "purchase")) %>%
  group_by(user_session, event_type) %>%
  summarise(
    event_count = n(),
    total_price = sum(price, na.rm = TRUE)
  ) %>%
  pivot_wider(names_from = event_type, values_from = c(event_count, total_price), values_fill = 0) %>%
  mutate(
    cart_to_purchase_ratio_count = ifelse(event_count_cart > 0, event_count_purchase / event_count_cart, NA),
    cart_to_purchase_ratio_price = ifelse(total_price_cart > 0, total_price_purchase / total_price_cart, NA)
  )

# View the results
cart_purchase_ratio_session
```

# **Step 4: Isolate baseline month data and create month in which change occurs**

## Step 4a: Extract baseline (February) data and create March (change) data

### Extract baseline month data
Here, we are using the latest month of data, February
```{r}
# Filter for baseline month (February) and store in a new variable
month_baseline <- deduplicateddf %>%
  filter(time_month == 2)

# View the filtered dataset
head(month_baseline)
summary(month_baseline)
```

### Create March data
```{r}
month_3 <- month_baseline
month_3$time_month <- 3

head(month_3)
```

## Step 4b: Determine March purchase to cart ratio and alter March data to fit specific problem/challenge

### Determine March purchase to cart ratio
```{r}
initial_cart_count <- nrow(filter(month_3, event_type == "cart"))
initial_purchase_count <- nrow(filter(month_3, event_type == "purchase"))
initial_ratio <- initial_purchase_count / initial_cart_count
initial_ratio
```

### Apply a reduction to purchase events in March
```{r}
# Calculate reduction of the purchase events
num_to_remove <- round(initial_purchase_count * 0.15)

# Randomly sample and remove above percentage of purchase events
set.seed(123)  # For reproducibility
purchases_to_remove <- sample(which(month_3$event_type == "purchase"), num_to_remove)
month_3_modified <- month_3[-purchases_to_remove, ]
```

### Calculate final cart-to-purchase ratio for March
```{r}
final_cart_count <- nrow(filter(month_3_modified, event_type == "cart"))
final_purchase_count <- nrow(filter(month_3_modified, event_type == "purchase"))
final_ratio <- final_purchase_count / final_cart_count
final_ratio
cat("Final Cart-to-Purchase Ratio for March after 10% reduction from February:", final_ratio, "\n")
initial_ratio-final_ratio
```

## Step 4c: Combine baseline and modified March data
```{r}
data_combined <- bind_rows(month_baseline, month_3_modified)

head(data_combined)
summary(data_combined)
```

# **EXPLORATORY DATA ANALYSIS**

# Step 5: Exploratory analysis on change month and differential from baseline month

## Step 5a: Report lost revenue from February to March

- The total loss in revenue is $213,921.06 
- The mean purchase value declined from $180.16 to $179.02

### Identify cumulative and mean revenue difference by event type
```{r}
#Baseline month
actionSumsBaseline <- tapply(month_baseline$price, month_baseline$event_type, sum)
actionSums_dollarBaseline <- dollar(actionSumsBaseline)
actionSums_dollarBaseline

actionMeanBaseline <- tapply(month_baseline$price, month_baseline$event_type, mean)
actionMeans_dollarBaseline <- dollar(actionMeanBaseline)
actionMeans_dollarBaseline

#Month 3/March
actionSumsMarch <- tapply(month_3_modified$price, month_3_modified$event_type, sum)
actionSums_dollarMarch <- dollar(actionSumsMarch)
actionSums_dollarMarch

actionMeanMarch <- tapply(month_3_modified$price, month_3_modified$event_type, mean)
actionMeans_dollarMarch <- dollar(actionMeanMarch)
actionMeans_dollarMarch
```

### Determine total loss in price/value from baseline to March
```{r}
#Total value loss
actionSums_dollarBaselinePurchase <- month_baseline %>%
  filter(event_type == "purchase") %>%
  summarise(total_revenue = sum(price))

print(actionSums_dollarBaselinePurchase)

actionSums_dollarMarchPurchase <- month_3_modified %>%
  filter(event_type == "purchase") %>%
  summarise(total_revenue = sum(price))

print(actionSums_dollarMarchPurchase)

lost_revenue <- actionSums_dollarBaselinePurchase-actionSums_dollarMarchPurchase
message <- paste("The estimated lost monthly revenue is", lost_revenue)
message
```

# Step 5b: Factor Analysis

In this section, we will examine how revenue was impacted in terms of several factors: products, pricing, and user engagement features.

### Products

What were the specific products most impacted by the change in terms of lost revenue? 
```{r}
# Filter only purchases
baseline_purchases <- month_baseline %>% filter(event_type == "purchase")
month3_purchases <- month_3_modified %>% filter(event_type == "purchase")

# Summarize revenue by brand and product categories
baseline_revenue <- baseline_purchases %>%
  group_by(brand, category_1, category_2, category_3) %>%
  summarise(total_revenue_baseline = sum(price, na.rm = TRUE), .groups = "drop")

month3_revenue <- month3_purchases %>%
  group_by(brand, category_1, category_2, category_3) %>%
  summarise(total_revenue_month3 = sum(price, na.rm = TRUE), .groups = "drop")

# Merge datasets by brand and product categories
revenue_comparison <- merge(
  baseline_revenue, 
  month3_revenue, 
  by = c("brand", "category_1", "category_2", "category_3"), 
  all = TRUE
)

# Replace NA with 0 (if a brand-category combination is missing in a month)
revenue_comparison[is.na(revenue_comparison)] <- 0

# Calculate revenue decline
revenue_comparison <- revenue_comparison %>%
  mutate(revenue_decline = total_revenue_baseline - total_revenue_month3)

# Sort by largest declines
largest_declines <- revenue_comparison %>%
  arrange(desc(revenue_decline))

# View the table with the biggest declines
print(head(largest_declines, 50))
```


```{r}
# Best sellers for March
best_sellers_march <- revenue_comparison %>%
  arrange(desc(total_revenue_month3))  # Sort by revenue for March

# View the top 50 best sellers in March
print(head(best_sellers_march, 10))
```

How were product brands and categories impacted overall by the change in terms of lost revenue?
```{r}
# Summarize revenue declines by brand
brand_declines <- revenue_comparison %>%
  group_by(brand) %>%
  summarise(
    total_revenue_baseline = sum(total_revenue_baseline, na.rm = TRUE),
    total_revenue_month3 = sum(total_revenue_month3, na.rm = TRUE)
  ) %>%
  mutate(revenue_decline = total_revenue_baseline - total_revenue_month3) %>%
  arrange(desc(revenue_decline))

# Summarize revenue declines by category_1
category_1_declines <- revenue_comparison %>%
  group_by(category_1) %>%
  summarise(
    total_revenue_baseline = sum(total_revenue_baseline, na.rm = TRUE),
    total_revenue_month3 = sum(total_revenue_month3, na.rm = TRUE)
  ) %>%
  mutate(revenue_decline = total_revenue_baseline - total_revenue_month3) %>%
  arrange(desc(revenue_decline))

# Summarize revenue declines by category_2
category_2_declines <- revenue_comparison %>%
  group_by(category_2) %>%
  summarise(
    total_revenue_baseline = sum(total_revenue_baseline, na.rm = TRUE),
    total_revenue_month3 = sum(total_revenue_month3, na.rm = TRUE)
  ) %>%
  mutate(revenue_decline = total_revenue_baseline - total_revenue_month3) %>%
  arrange(desc(revenue_decline))

# Summarize revenue declines by category_3
category_3_declines <- revenue_comparison %>%
  group_by(category_3) %>%
  summarise(
    total_revenue_baseline = sum(total_revenue_baseline, na.rm = TRUE),
    total_revenue_month3 = sum(total_revenue_month3, na.rm = TRUE)
  ) %>%
  mutate(revenue_decline = total_revenue_baseline - total_revenue_month3) %>%
  arrange(desc(revenue_decline))

# View top declines for each summary
print("Top Brand Declines:")
print(head(brand_declines, 10))

print("Top Category_1 Declines:")
print(head(category_1_declines, 10))

print("Top Category_2 Declines:")
print(head(category_2_declines, 10))

print("Top Category_3 Declines:")
print(head(category_3_declines, 10))
```

### Pricing
Which price points were most impacted by the change in terms of lost revenue? 
```{r}
# Add price bucket to the original datasets
bucket_size <- 100  

baseline_purchases <- baseline_purchases %>%
  mutate(price_bucket = floor(price / bucket_size) * bucket_size)

month3_purchases <- month3_purchases %>%
  mutate(price_bucket = floor(price / bucket_size) * bucket_size)

# Summarize revenue by price bucket for baseline and month 3
baseline_revenue <- baseline_purchases %>%
  group_by(price_bucket) %>%
  summarise(total_revenue_baseline = sum(price, na.rm = TRUE), .groups = "drop")

month3_revenue <- month3_purchases %>%
  group_by(price_bucket) %>%
  summarise(total_revenue_month3 = sum(price, na.rm = TRUE), .groups = "drop")

# Merge datasets to compare revenues
price_revenue_comparison <- merge(
  baseline_revenue, 
  month3_revenue, 
  by = "price_bucket", 
  all = TRUE
)

# Replace NA values with 0 (in case some price buckets are missing in one month)
price_revenue_comparison[is.na(price_revenue_comparison)] <- 0

# Calculate revenue decline for each price bucket
price_revenue_comparison <- price_revenue_comparison %>%
  mutate(revenue_decline = total_revenue_baseline - total_revenue_month3) %>%
  arrange(desc(revenue_decline))

# View top price buckets with declines
print("Top Price Buckets with Declines:")
print(head(price_revenue_comparison, 10))
```

Visualize the relationship between price and revenue decline
```{r}
# Plot revenue declines by price buckets
ggplot(price_revenue_comparison, aes(x = price_bucket, y = revenue_decline)) +
  geom_bar(stat = "identity", fill = "steelblue", color = "black") +
  labs(
    title = "Revenue Declines by Price Buckets",
    x = "Price Bucket",
    y = "Revenue Decline",
    caption = "Data shows the difference between baseline and month 3 revenue"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5, size = 16, face = "bold"),
    axis.title = element_text(size = 12),
    axis.text = element_text(size = 10)
  ) +
  scale_x_continuous(breaks = seq(min(price_revenue_comparison$price_bucket, na.rm = TRUE),
                                  max(price_revenue_comparison$price_bucket, na.rm = TRUE),
                                  by = 100))  # Adjust tick marks as needed

# Calculate the correlation coefficient
correlation_value_price <- cor(price_revenue_comparison$price_bucket, price_revenue_comparison$revenue_decline, use = "complete.obs")

# Scatter plot with correlation coefficient
ggplot(price_revenue_comparison, aes(x = price_bucket, y = revenue_decline)) +
  geom_point(color = "blue", alpha = 0.6) + # Scatter points
  geom_smooth(method = "lm", se = FALSE, color = "red") + # Add a regression line
  labs(
    title = "Relationship Between Price Bucket and Revenue Decline",
    x = "Price Bucket",
    y = "Revenue Decline",
    caption = paste("Correlation coefficient: ", round(correlation_value_price, 2))
  ) +
  theme_minimal()

```

### User engagement
Which specific session engagement times were most impacted?
```{r}
# Filter only purchases from baseline and month 3
baseline_purchases <- month_baseline %>% filter(event_type == "purchase")
month3_purchases <- month_3_modified %>% filter(event_type == "purchase")

# Summarize revenue by time_day and time_hour
baseline_revenue <- baseline_purchases %>%
  group_by(time_day, time_hour) %>%
  summarise(total_revenue_baseline_time = sum(price, na.rm = TRUE), .groups = "drop")

month3_revenue <- month3_purchases %>%
  group_by(time_day, time_hour) %>%
  summarise(total_revenue_month3_time = sum(price, na.rm = TRUE), .groups = "drop")

# Merge datasets by time_day and time_hour
revenue_comparison <- merge(
  baseline_revenue, 
  month3_revenue, 
  by = c("time_day", "time_hour"), 
  all = TRUE
)

# Replace NA with 0 (if a time_day-time_hour combination is missing in a month)
revenue_comparison[is.na(revenue_comparison)] <- 0

# Calculate revenue decline
revenue_comparison <- revenue_comparison %>%
  mutate(revenue_decline_time = total_revenue_baseline_time - total_revenue_month3_time)

# Sort by largest declines
largest_declines <- revenue_comparison %>%
  arrange(desc(revenue_decline_time))

# View the table with the biggest declines
print(head(largest_declines, 50))
```

Which specific days or times of the week were most impacted?
```{r}
# Summarize revenue declines by time_day
time_day_declines <- revenue_comparison %>%
  group_by(time_day) %>%
  summarise(
    total_revenue_baseline_time = sum(total_revenue_baseline_time, na.rm = TRUE),
    total_revenue_month3_time = sum(total_revenue_month3_time, na.rm = TRUE)
  ) %>%
  mutate(revenue_decline_time = total_revenue_baseline_time - total_revenue_month3_time) %>%
  arrange(desc(revenue_decline_time))

# Summarize revenue declines by time_hour
time_hour_declines <- revenue_comparison %>%
  group_by(time_hour) %>%
  summarise(
    total_revenue_baseline_time = sum(total_revenue_baseline_time, na.rm = TRUE),
    total_revenue_month3_time = sum(total_revenue_month3_time, na.rm = TRUE)
  ) %>%
  mutate(revenue_decline_time = total_revenue_baseline_time - total_revenue_month3_time) %>%
  arrange(desc(revenue_decline_time))

# View top declines for each summary
print("Top Time Day Declines:")
print(head(time_day_declines, 10))

print("Top Time Hour Declines:")
print(head(time_hour_declines, 10))
```

```{r}
# Ensure event_time is in datetime format
month_baseline$event_time <- as.POSIXct(month_baseline$event_time, format="%Y-%m-%d %H:%M:%S", tz="UTC")
month_3_modified$event_time <- as.POSIXct(month_3_modified$event_time, format="%Y-%m-%d %H:%M:%S", tz="UTC")

# Calculate the session duration for each session in baseline data
session_duration_baseline <- month_baseline %>%
  group_by(user_session) %>%
  summarise(session_start = min(event_time), 
            session_end = max(event_time),
            session_duration = as.numeric(difftime(session_end, session_start, units = "secs")),
            .groups = "drop")

# Calculate the session duration for each session in month 3 data
session_duration_month3 <- month_3_modified %>%
  group_by(user_session) %>%
  summarise(session_start = min(event_time), 
            session_end = max(event_time),
            session_duration = as.numeric(difftime(session_end, session_start, units = "secs")),
            .groups = "drop")

# Summarize revenue by user session in baseline data
baseline_revenue_session <- month_baseline %>%
  group_by(user_session) %>%
  summarise(total_revenue_baseline = sum(price, na.rm = TRUE), .groups = "drop")

# Summarize revenue by user session in month 3 data
month3_revenue_session <- month_3_modified %>%
  group_by(user_session) %>%
  summarise(total_revenue_month3 = sum(price, na.rm = TRUE), .groups = "drop")

# Merge session duration and revenue data for both baseline and month3
session_revenue_comparison <- session_duration_baseline %>%
  left_join(session_duration_month3, by = "user_session") %>%
  left_join(baseline_revenue_session, by = "user_session") %>%
  left_join(month3_revenue_session, by = "user_session")

# Replace NA values in numeric columns with 0 (but not in user_session)
session_revenue_comparison <- session_revenue_comparison %>%
  mutate(across(where(is.numeric), ~ replace(., is.na(.), 0)))

# Calculate revenue decline for each session
session_revenue_comparison <- session_revenue_comparison %>%
  mutate(revenue_decline = total_revenue_baseline - total_revenue_month3)

# Sort session_revenue_comparison by revenue_decline in descending order
session_revenue_comparison_sorted <- session_revenue_comparison %>%
  arrange(desc(revenue_decline))

# View the top rows of the sorted data
head(session_revenue_comparison_sorted)
```

Visualize the relationship between session duration (month 3) and revenue decline
```{r}
# Calculate the correlation coefficient
correlation_value <- cor(session_revenue_comparison_sorted$session_duration.y, session_revenue_comparison_sorted$revenue_decline, use = "complete.obs")

# Scatter plot with correlation coefficient
ggplot(session_revenue_comparison_sorted, aes(x = session_duration.y, y = revenue_decline)) +
  geom_point(color = "blue", alpha = 0.6) + # Scatter points
  geom_smooth(method = "lm", se = FALSE, color = "red") + # Add a regression line
  labs(
    title = "Relationship Between Session Duration and Revenue Decline",
    x = "Session Duration (in seconds)",
    y = "Revenue Decline",
    caption = paste("Correlation coefficient: ", round(correlation_value, 2))
  ) +
  theme_minimal()
```

## Step 5c: User Engagement Feature Engineering

### Interaction counts
```{r}
# Count the number of views, carts, and purchases for each user session
interaction_counts <- month_3_modified %>%
  group_by(user_session) %>%
  summarize(
    views_count = sum(event_type == "view"),
    cart_count = sum(event_type == "cart"),
    purchase_count = sum(event_type == "purchase"),
    total_interactions = views_count + cart_count + purchase_count
  )

interaction_counts_long <- interaction_counts %>%
  pivot_longer(cols = c(views_count, cart_count, purchase_count), names_to = "interaction_type", values_to = "count")

ggplot(interaction_counts_long, aes(x = interaction_type, y = count, fill = interaction_type)) +
  geom_boxplot(outlier.colour = "red", outlier.shape = 1, outlier.size = 2) +
  labs(title = "Boxplot of Interaction Counts", x = "Interaction Type", y = "Count") +
  theme_minimal() +
  theme(legend.position = "none")

```

The boxplot illustrates the distribution of interaction counts (views, cart additions, and purchases) across sessions, highlighting significant engagement in product views compared to relatively lower cart additions and purchases.

### Session duration
```{r}
# Convert event_time to POSIXct
month_3_modified$event_time <- as.POSIXct(month_3_modified$event_time, format = "%Y-%m-%d %H:%M:%S", tz = "UTC")

# Calculate session duration per user session
session_duration <- month_3_modified %>%
  group_by(user_session) %>%
  summarize(session_duration = as.numeric(difftime(max(event_time), min(event_time), units = "secs")))

```

### User interaction frequency
```{r}
# Combine total interactions and session duration to calculate interaction frequency
interaction_frequency <- interaction_counts %>%
  left_join(session_duration, by = "user_session") %>%
  mutate(interaction_frequency = total_interactions / (session_duration / 60 + 1))  # Avoid division by zero

```

### Category diversity
```{r}
# Count the number of unique categories per session
category_diversity <- month_3_modified %>%
  group_by(user_session) %>%
  summarize(category_diversity = n_distinct(category_1))

```

### Merge features
```{r}
# Merge all features into the main dataset
features <- interaction_counts %>%
  left_join(session_duration, by = "user_session") %>%
  left_join(interaction_frequency %>% select(user_session, interaction_frequency), by = "user_session") %>%
  left_join(category_diversity, by = "user_session")

# Add features to the main dataset
month_3_modified <- month_3_modified %>%
  left_join(features, by = "user_session")
```

```{r}
# Create a binary field indicating if a session resulted in a conversion from cart to purchase
cart_to_purchase <- month_3_modified %>%
  group_by(user_session) %>%
  summarize(
    cart_events = sum(event_type == "cart"),  # Count cart events
    purchase_events = sum(event_type == "purchase"),  # Count purchase events
    cart_to_purchase = ifelse(cart_events > 0 & purchase_events > 0, 1, 0)  # 1 if conversion occurred, 0 otherwise
  )

# Merge cart-to-purchase conversion field into the main dataset
month_3_modified <- month_3_modified %>%
  left_join(cart_to_purchase, by = "user_session")
```

# **PREDICTIVE MODELING**

In this section, our goal is to understand the features and predict the decision likelihood for cart-to-purchase conversion 

We will use classification tree, random forest, and logistic regression and compare the performance of each model.

# **Step 6: Plot and compare all three models**

## Step 6a: Prepare dataset
```{r}
# Check for missing values in the dataset
colSums(is.na(month_3_modified))

# Remove rows with any missing values
month_3_modified <- month_3_modified %>%
  drop_na()
```

## Step 6b: Training and test sets
```{r}
# Ensure cart_to_purchase is a factor
month_3_modified$cart_to_purchase <- as.factor(month_3_modified$cart_to_purchase)

# Create training and validation sets
set.seed(123)  # For reproducibility
train_index <- createDataPartition(month_3_modified$cart_to_purchase, p = 0.8, list = FALSE)
train_data <- month_3_modified[train_index, ]
test_data <- month_3_modified[-train_index, ]
```

## Step 6c: Compare random forest, classification tree, and logistic models
```{r}
#Convert Y variable to factor type
train_data$cart_to_purchase <- as.factor(train_data$cart_to_purchase)
test_data$cart_to_purchase <- as.factor(test_data$cart_to_purchase)

# Random Forest Model
rf_model <- randomForest(
  formula = cart_to_purchase ~ views_count + cart_count + session_duration + price + category_diversity + category_1,
  data = train_data,
  ntree = 400,
  replace = TRUE,
  importance = TRUE
)

# Logistic Regression model
logistic_model <- glm(cart_to_purchase ~ views_count + cart_count + session_duration + price + category_diversity + category_1, 
                      family = "binomial", data = train_data)

# Classification Tree model
tree_model <- rpart(cart_to_purchase ~ views_count + cart_count + session_duration + price + category_diversity + category_1, 
                    method = "class", data = train_data)
printcp(tree_model)

# Random Forest model predictions
pred.prob.rf <- predict(rf_model, test_data, type = "prob")[, 2]

# Logistic model predictions
pred.prob.logistic <- predict(logistic_model, test_data, type = "response")

# Pruned classification tree predictions
pred.prob.tree <- predict(tree_model, test_data, type = "prob")[, 2]

# ROC for Random Forest, Logistic, and Tree models
roc.rf <- performance(prediction(pred.prob.rf, test_data$cart_to_purchase), "sens", "fpr")
roc.logistic <- performance(prediction(pred.prob.logistic, test_data$cart_to_purchase), "sens", "fpr")
roc.tree <- performance(prediction(pred.prob.tree, test_data$cart_to_purchase), "sens", "fpr")

# Plot all three ROC curves
plot(roc.rf)
plot(roc.logistic, col = "red", add = TRUE)
plot(roc.tree, col = "blue", add = TRUE)

# Sensitivity chart for all models
plot(roc.rf, xlim = c(0.01, 0.4), ylim = c(0.5, 0.96))
plot(roc.logistic, col = "red", add = TRUE)
plot(roc.tree, col = "blue", add = TRUE)

# LIFT curve for all models
lift.rf <- performance(prediction(pred.prob.rf, test_data$cart_to_purchase), "lift", "rpp")
lift.logistic <- performance(prediction(pred.prob.logistic, test_data$cart_to_purchase), "lift", "rpp")
lift.tree <- performance(prediction(pred.prob.tree, test_data$cart_to_purchase), "lift", "rpp")

# Plot all three LIFT curves
plot(lift.rf)
plot(lift.logistic, col = "red", add = TRUE)
plot(lift.tree, col = "blue", add = TRUE)
```
```{r}
# AUC for all models
auc.rf <- performance(prediction(pred.prob.rf, test_data$cart_to_purchase), "auc")@y.values
auc.logistic <- performance(prediction(pred.prob.logistic, test_data$cart_to_purchase), "auc")@y.values
auc.tree <- performance(prediction(pred.prob.tree, test_data$cart_to_purchase), "auc")@y.values

auc.rf
auc.logistic
auc.tree
```

Based on the higher area under the curve for random forest, we will select this as our conversion predictive model.

## Step 6d: Analyze the Random Forest Model
```{r}
# Plot the Random Forest model
plot(rf_model)

# Plot the variable importance
varImpPlot(rf_model)

# Predict the probabilities of the second class (1 = cart-to-purchase)
pred.prob.rf <- predict(rf_model, test_data, type = "prob")[, 2]

# Create confusion matrix
conf_matrix <- table(test_data$cart_to_purchase, pred.prob.rf > 0.5)
conf_matrix

# Extract values from the confusion matrix
true_negatives <- conf_matrix[1, 1]  # Actual 0, Predicted FALSE
false_positives <- conf_matrix[1, 2]  # Actual 0, Predicted TRUE
false_negatives <- conf_matrix[2, 1]  # Actual 1, Predicted FALSE
true_positives <- conf_matrix[2, 2]  # Actual 1, Predicted TRUE

true_negatives
false_positives
false_negatives
true_positives

# Misclassification rate
misclassification_rate <- (false_positives + false_negatives) / sum(conf_matrix)
misclassification_rate

# Class 1 error rate (False Negative Rate)
class_1_error_rate <- false_negatives / (true_positives + false_negatives)
class_1_error_rate

# Class 0 error rate (False Positive Rate)
class_0_error_rate <- false_positives / (false_positives + true_negatives)
class_0_error_rate

# Precision: True Positives / (True Positives + False Positives)
precision <- true_positives / (true_positives + false_positives)
precision

# Negative Predictive Value: True Negatives / (True Negatives + False Negatives)
npv <- true_negatives / (true_negatives + false_negatives)
npv

# ROC and AUC
pred.rf <- prediction(pred.prob.rf, test_data$cart_to_purchase)
performance(pred.rf, measure = "auc")@y.values
roc.rf <- performance(pred.rf, "sens", "fpr")
plot(roc.rf)

# LIFT curve
lift.rf <- performance(pred.rf, "lift", "rpp")
plot(lift.rf)

# Sensitivity chart
plot(roc.rf, xlim = c(0.05, 0.10), ylim = c(0.96, 0.99))
```

# **Step 7: Analysis to recover lost revenue - overall increased conversion model**
Here, we provide the user with a tool to simulate a third month in which the higher likelihood sessions are promoted, sufficient to ensure that the total value of price column for purchases increases from $1,162,885 in March to $1,376,806 in April.

## Step 7a: Apply prediction likelihood to dataset
```{r}
# Ensure cart_to_purchase is a factor for prediction
month_3_modified$cart_to_purchase <- as.factor(month_3_modified$cart_to_purchase)

# Predict the probabilities for the entire month_3_modified dataset
pred.prob.rf <- predict(rf_model, month_3_modified, type = "prob")[, 2]

# Add the predicted probabilities to the dataset as conversion likelihood
month_3_modified$conversion_likelihood <- pred.prob.rf

# Check the first few rows to ensure the likelihood is correctly added
head(month_3_modified)
```

## Step 7b: Simulate revenue recovery
```{r}
#Step 1: Define the top x% of purchase likelihood sessions to replicate and bottom x% to remove
top_percentage <- 0.21  # Example: Promote top 20% of sessions
bottom_percentage <- 0.2  # Example: Remove bottom 20% of sessions

```

```{r}
#Step 2: Filter purchase events
purchase_data <- month_3_modified %>% filter(event_type == "purchase")
```

```{r}
#Step 3: Sort purchase data by conversion likelihood (highest to lowest)
purchase_data_sorted <- purchase_data %>%
  arrange(desc(conversion_likelihood))
head(purchase_data_sorted)
```

```{r}
#Step 4: Calculate how many rows to keep for top x% and bottom x%
top_rows_count <- floor(nrow(purchase_data_sorted) * top_percentage)
bottom_rows_count <- floor(nrow(purchase_data_sorted) * bottom_percentage)
```

```{r}
#Step 5: Select top x% and bottom x% sessions
top_sessions <- purchase_data_sorted[1:top_rows_count, ]
bottom_sessions <- purchase_data_sorted[(nrow(purchase_data_sorted) - bottom_rows_count + 1):nrow(purchase_data_sorted), ]
```

```{r}
#Step 6: Remove the bottom x% sessions from purchase data
purchase_data_modified <- purchase_data_sorted[!purchase_data_sorted$user_session %in% bottom_sessions$user_session, ]
```

```{r}
#Step 7: Replicate the top x% sessions
purchase_data_modified <- rbind(purchase_data_modified, top_sessions)
```

```{r}
#Step 8: Combine the modified purchase data with the non-purchase data
non_purchase_data <- month_3_modified %>% filter(event_type != "purchase")
month_3_modified_replicated <- rbind(purchase_data_modified, non_purchase_data)
```

```{r}
#Step 9: Calculate the total purchase value for the replicated dataset before and after modification
total_purchase_value_original <- sum(month_3_modified$price[month_3_modified$event_type == "purchase"])
total_purchase_value_modified <- sum(month_3_modified_replicated$price[month_3_modified_replicated$event_type == "purchase"])
```

```{r}
 #Step 10: Ensure the purchase value for Month 4 matches the target value
target_total_value_month4 <- 1376806  # Target purchase value for Month 4

if (total_purchase_value_modified < target_total_value_month4) {
   #Calculate how much value we still need to add
  value_needed <- target_total_value_month4 - total_purchase_value_modified
  
  #If necessary, replicate the top sessions further to reach the target value
  additional_sessions_needed <- ceiling(value_needed / mean(top_sessions$price))
  additional_sessions <- top_sessions[sample(nrow(top_sessions), additional_sessions_needed, replace = TRUE), ]
  
  #Add the additional sessions to the modified data
  month_3_modified_replicated <- rbind(month_3_modified_replicated, additional_sessions)
  
 # Recalculate the new total purchase value
  total_purchase_value_modified <- sum(month_3_modified_replicated$price[month_3_modified_replicated$event_type == "purchase"])
  
  cat("Adjusted purchase value for Month 4 (after adding more top sessions):", total_purchase_value_modified, "\n")
}
```

```{r}
#Step 11: Save the modified dataset
write.csv(month_3_modified_replicated, "month_4_simulated_with_adjustment.csv", row.names = FALSE)
cat("Month 4 simulated dataset saved as 'month_4_simulated_with_adjustment.csv'.\n")
```

## Step 7c: Summarize overall revenue recovery of increased likelihood model

- **Original month 2 total purchase value:** 1,376,806
- **Month 3 with 10% decline total purchase value:** 1,162,885
- **Month 4 modified total purchase value:** 1,379,896

```{r}
#Revenue increase from March to April
1379896-1162885
```


## Step 7d: Analyze features of top sessions

### Summary of most impactful features
```{r}
#Show most impactful variables in random forest model
varImpPlot(rf_model)

#Analyze the features of the top sessions
top_sessions_features <- top_sessions %>%
  select(conversion_likelihood, views_count, cart_count, session_duration, price, category_diversity, category_1)

#Summarize the features of the top sessions
summary(top_sessions_features)
```

### Analysis of specific features

**Views Count**
```{r}
# Categorize views count into ranges
top_sessions_features <- top_sessions_features %>%
  mutate(
    views_range = cut(
      views_count,
      breaks = c(1, 5, 10, 20, 50, Inf),
      labels = c("1-4", "5-9", "10-19", "20-49", "50+"),
      include.lowest = TRUE
    )
  )

# Calculate mean likelihood and relative difference by views count range
views_summary <- top_sessions_features %>%
  group_by(views_range) %>%
  summarise(mean_likelihood = mean(conversion_likelihood, na.rm = TRUE)) %>%
  mutate(relative_difference = (mean_likelihood - mean(mean_likelihood)) / mean(mean_likelihood))

# Plot the relative difference by views count range
ggplot(views_summary, aes(x = views_range, y = relative_difference * 100, fill = views_range)) +
  geom_bar(stat = "identity") +
  labs(
    title = "Relative Difference in Conversion Likelihood by Views Count Range",
    x = "Views Count Range",
    y = "Relative Difference (%)"
  ) +
  theme_minimal() +
  theme(legend.position = "none")
```

**Cart Count**
```{r}
# Categorize cart count into ranges
top_sessions_features <- top_sessions_features %>%
  mutate(
    cart_range = cut(
      cart_count,
      breaks = c(1, 2, 3, 4, 5, 6, 7, 8, 9, 10, Inf),
      labels = c("1", "2", "3", "4", "5", "6", "7", "8", "9", "10+"),
      include.lowest = TRUE
    )
  )

# Calculate mean likelihood and relative difference by cart count range
cart_summary <- top_sessions_features %>%
  group_by(cart_range) %>%
  summarise(mean_likelihood = mean(conversion_likelihood, na.rm = TRUE)) %>%
  mutate(relative_difference = (mean_likelihood - mean(mean_likelihood)) / mean(mean_likelihood))

# Plot the relative difference by cart count range
ggplot(cart_summary, aes(x = cart_range, y = relative_difference * 100, fill = cart_range)) +
  geom_bar(stat = "identity") +
  labs(
    title = "Relative Difference in Conversion Likelihood by Cart Count Range",
    x = "Cart Count Range",
    y = "Relative Difference (%)"
  ) +
  theme_minimal() +
  theme(legend.position = "none")
```

**Session Duration**
```{r}
# Add session_duration_minutes and categorize into session ranges
top_sessions_features <- top_sessions_features %>%
  mutate(
    session_duration_minutes = session_duration / 60,  # Convert session duration to minutes
    session_range = cut(
      session_duration_minutes,
      breaks = c(1, 5, 10, 15, 30, 45, 60, Inf),
      labels = c("1-4", "5-9", "10-14", "15-29", "30-44", "45-59", "60+"),
      include.lowest = TRUE
    )
  )
 
# Calculate mean likelihood and relative difference by session duration range
session_summary <- top_sessions_features %>%
  group_by(session_range) %>%
  summarise(mean_likelihood = mean(conversion_likelihood, na.rm = TRUE)) %>%
  mutate(relative_difference = (mean_likelihood - mean(mean_likelihood)) / mean(mean_likelihood))
 
# Plot the relative difference by session duration range
ggplot(session_summary, aes(x = session_range, y = relative_difference * 100, fill = session_range)) +
  geom_bar(stat = "identity") +
  labs(
    title = "Relative Difference in Conversion Likelihood by Session Duration Range in Minutes",
    x = "Session Duration Range in Minutes",
    y = "Relative Difference (%)"
  ) +
  theme_minimal() +
  theme(legend.position = "none")
```

**Price Bucket**
```{r}
top_sessions_features <- top_sessions_features %>%
  mutate(
    price_range = cut(
      price,
      breaks = c(1, 100, 200, 300, 400, 500, 600, 700, 800, 900, Inf),
      labels = c("$1-$99", "$100-$199", "$200-$299", "$300-$399", "$400-$499", "$500-$599", "$600-$699", "$700-$799", "$800-$899", "$900+"),
      include.lowest = TRUE
    )
  )

# Calculate relative differences
overall_mean <- mean(top_sessions_features$conversion_likelihood, na.rm = TRUE)
price_range_summary <- top_sessions_features %>%
  group_by(price_range) %>%
  summarise(mean_likelihood = mean(conversion_likelihood, na.rm = TRUE)) %>%
  mutate(relative_difference = (mean_likelihood - overall_mean) / overall_mean * 100)
 
# Plot relative differences
ggplot(price_range_summary, aes(x = price_range, y = relative_difference, fill = price_range)) +
  geom_bar(stat = "identity") +
  labs(
    title = "Relative Difference in Conversion Likelihood by Price Range",
    x = "Price Range",
    y = "Relative Difference (%)"
  ) +
  theme_minimal() +
  theme(legend.position = "none")
```

**Category Diversity**
```{r}
# Compare single vs. multi-category users
top_sessions_features <- top_sessions_features %>%
  mutate(
    category_diversity_range = cut(
      category_diversity,
      breaks = c(1, 2, Inf),
      labels = c("Single Category", "Multi-Category"),
      include.lowest = TRUE
    )
  )

# Calculate mean likelihood and relative difference by category diversity range
category_diversity_summary <- top_sessions_features %>%
  group_by(category_diversity_range) %>%
  summarise(mean_likelihood = mean(conversion_likelihood, na.rm = TRUE)) %>%
  mutate(relative_difference = (mean_likelihood - mean(mean_likelihood)) / mean(mean_likelihood))

# Plot the relative difference by category diversity range
ggplot(category_diversity_summary, aes(x = category_diversity_range, y = relative_difference * 100, fill = category_diversity_range)) +
  geom_bar(stat = "identity") +
  labs(
    title = "Relative Difference in Conversion Likelihood by Category Diversity Range",
    x = "Category Diversity Range",
    y = "Relative Difference (%)"
  ) +
  theme_minimal() +
  theme(legend.position = "none")

# Categorize category diversity into ranges
top_sessions_features <- top_sessions_features %>%
  mutate(
    category_diversity_range = cut(
      category_diversity,
      breaks = c(1, 1.5, 2, 2.5, 3, 3.5, 4, 4.5, 5, Inf),
      labels = c("1-1.5", "1.5-2", "2-2.5", "2.5-3", "3-3.5", "3.5-4", "4-4.5", "4.5-5", "5+"),
      include.lowest = TRUE
    )
  )

# Calculate mean likelihood and relative difference by category diversity range
category_diversity_summary <- top_sessions_features %>%
  group_by(category_diversity_range) %>%
  summarise(mean_likelihood = mean(conversion_likelihood, na.rm = TRUE)) %>%
  mutate(relative_difference = (mean_likelihood - mean(mean_likelihood)) / mean(mean_likelihood))

# Plot the relative difference by category diversity range
ggplot(category_diversity_summary, aes(x = category_diversity_range, y = relative_difference * 100, fill = category_diversity_range)) +
  geom_bar(stat = "identity") +
  labs(
    title = "Relative Difference in Conversion Likelihood by Category Diversity Range",
    x = "Category Diversity Range",
    y = "Relative Difference (%)"
  ) +
  theme_minimal() +
  theme(legend.position = "none")
```

**High Level Category**
```{r}
# Calculate mean likelihood and relative difference for each category in category_1
category_1_summary <- top_sessions_features %>%
  group_by(category_1) %>%
  summarise(mean_likelihood = mean(conversion_likelihood, na.rm = TRUE)) %>%
  mutate(relative_difference = (mean_likelihood - mean(mean_likelihood)) / mean(mean_likelihood))

# Plot the relative difference for each category in category_1
ggplot(category_1_summary, aes(x = reorder(category_1, relative_difference), y = relative_difference * 100, fill = category_1)) +
  geom_bar(stat = "identity") +
  labs(
    title = "Relative Difference in Conversion Likelihood by Category",
    x = "Category",
    y = "Relative Difference (%)"
  ) +
  theme_minimal() +
  theme(legend.position = "none") +
  coord_flip()  # Flip the coordinates for better readability if there are many categories
```

# **Step 8: Linear Regression for Single Variable**

Here, we are testing how increasing the value of a single variable (such as the number of items in a cart) influence purchase revenue.

## Step 8a: Summarize data and model performance
```{r}
summary(month_3_modified_replicated)
```

The linear regression model using total revenue as the response variable explains 33.41% of the variance (R² = 0.3341) and has the lowest residual standard error (288), outperforming both the log-transformed revenue model (R² = 0.05943, residual standard error = 1.164) and the square root-transformed revenue model (R² = 0.1751, residual standard error = 4.991) in terms of explained variance and predictive accuracy.

## Step 8b: Create and compare linear models
```{r}
library(dplyr)
library(ggplot2)

# Step 1: Aggregate and transform data
session_data <- month_3_modified_replicated %>%
  group_by(user_session) %>%
  summarise(
    total_cart_additions = sum(cart_count, na.rm = TRUE),
    total_revenue = sum(price * (event_type == "purchase"), na.rm = TRUE),
    total_purchases = sum(event_type == "purchase", na.rm = TRUE)
  ) %>%
  mutate(
    revenue_log = ifelse(total_revenue > 0, log(total_revenue), 0),
    revenue_sqrt = sqrt(total_revenue),
    cart_to_purchase_ratio = ifelse(total_cart_additions > 0, total_purchases / total_cart_additions, 0)
  )

# Step 2: Fit Linear Models
model_revenue <- lm(total_revenue ~ total_cart_additions, data = session_data)
model_revenue_log <- lm(revenue_log ~ total_cart_additions, data = session_data)
model_revenue_sqrt <- lm(revenue_sqrt ~ total_cart_additions, data = session_data)

# Step 3: Summarize Models
summary_revenue <- summary(model_revenue)
summary_revenue_log <- summary(model_revenue_log)
summary_revenue_sqrt <- summary(model_revenue_sqrt)

# Print Summaries
cat("Summary for Revenue Model:\n")
print(summary_revenue)

cat("\nSummary for Log Revenue Model:\n")
print(summary_revenue_log)

cat("\nSummary for Sqrt Revenue Model:\n")
print(summary_revenue_sqrt)

# Step 4: Compare Metrics
r_squared <- c(
  summary_revenue$r.squared,
  summary_revenue_log$r.squared,
  summary_revenue_sqrt$r.squared
)

rmse <- c(
  sqrt(mean(model_revenue$residuals^2)),
  sqrt(mean(model_revenue_log$residuals^2)),
  sqrt(mean(model_revenue_sqrt$residuals^2))
)

model_comparison <- data.frame(
  Model = c("Revenue", "Log Revenue", "Sqrt Revenue"),
  R_Squared = r_squared,
  RMSE = rmse
)

# Step 5: Print Model Comparison
cat("\nModel Comparison:\n")
print(model_comparison)

# Step 6: Visualize Linear Relationships
ggplot(session_data, aes(x = total_cart_additions, y = total_revenue)) +
  geom_point() +
  geom_smooth(method = "lm", color = "blue") +
  labs(
    title = "Linear Regression: Total Cart Additions vs. Revenue",
    x = "Total Cart Additions",
    y = "Revenue"
  ) +
  theme_minimal()

ggplot(session_data, aes(x = total_cart_additions, y = revenue_log)) +
  geom_point() +
  geom_smooth(method = "lm", color = "blue") +
  labs(
    title = "Linear Regression: Total Cart Additions vs. Log Revenue",
    x = "Total Cart Additions",
    y = "Log Revenue"
  ) +
  theme_minimal()

ggplot(session_data, aes(x = total_cart_additions, y = revenue_sqrt)) +
  geom_point() +
  geom_smooth(method = "lm", color = "blue") +
  labs(
    title = "Linear Regression: Total Cart Additions vs. Sqrt Revenue",
    x = "Total Cart Additions",
    y = "Sqrt Revenue"
  ) +
  theme_minimal()


plot(model_revenue)

```
```{r}
# Calculate Cook's Distance
cooks_distance <- cooks.distance(model_revenue)

# Add Cook's Distance to the dataset for analysis
session_data$cooks_distance <- cooks_distance

# Threshold for identifying influential points
threshold <- 4 / nrow(session_data)

# Identify high-leverage points
high_leverage_points <- which(cooks_distance > threshold)

# Display high-leverage points
cat("Number of high-leverage points:", length(high_leverage_points), "\n")
cat("High-leverage points indices:\n")
print(high_leverage_points)

# View the corresponding rows in the dataset
high_leverage_data <- session_data[high_leverage_points, ]
head(high_leverage_data)
```

```{r}
# Remove high-leverage points from the dataset
session_data_filtered <- session_data[-high_leverage_points, ]

# Refit the model with the filtered data
model_revenue_filtered <- lm(total_revenue ~ total_cart_additions, data = session_data_filtered)

# Summary of the updated model
summary(model_revenue_filtered)

# Diagnostic plots for the new model
plot(model_revenue_filtered)
```

## Step 8c: Predict revenue impact
```{r}
# Step 7: Predict the impact of a 20% increase in cart additions
current_cart_additions <- sum(session_data$total_cart_additions, na.rm = TRUE)
current_revenue <- coef(model_revenue)[1] + coef(model_revenue)[2] * current_cart_additions
increased_cart_additions <- current_cart_additions * 1.2
increased_revenue <- coef(model_revenue)[1] + coef(model_revenue)[2] * increased_cart_additions

impact <- increased_revenue - current_revenue

cat("\nCurrent Revenue :", current_revenue, "\n")
cat("Increased Revenue :", increased_revenue, "\n")
cat("Impact of 20% Increase in Cart Additions:", impact, "\n")
```

A 20% increase in cart additions could increase current revenue from $619,940 to $743,924, resulting in an additional $123,984 in revenue. However, this projection relies on a regression model with limited robustness due to low R-squared values, potential non-linearity in relationships, and heteroscedasticity observed in the residuals, indicating that further refinement of the model is needed for more accurate predictions.

# **Recommended Solutions**

## **Overview**
The following strategies for recovering lost revenue are based on our analysis, including results from the random forest model. This model was the most successful at identifying the key factors driving cart-to-purchase conversions, allowing us to recommend targeted actions to influence conversion that are grounded in statistical insights.

---

## **Key Findings**
The analysis identified the following variables as the most significant predictors of cart-to-purchase conversions:

1. **Cart Count**: Users adding more items to their cart are significantly more likely to convert.
2. **Session Duration**: Longer session times are associated with higher conversion rates.
3. **Views Count**: Engaged users who explore more products are more likely to make a purchase.
4. **Price**: Low to mid-range priced items (<$300) are most likely to convert, while higher-priced items show a drop in conversions.
5. **Category Diversity**: Users interacting with multiple categories of products are more likely to purchase.
6. **Category_1**: Specific high-level product categories influence the likelihood of conversion.

## **Recommended Strategies**

### **1. Maximize Cart Additions**
- **Why?** Modeling results show that cart count is the strongest driver of conversion.
- **Action Steps:**
  - Encourage users to add items to their cart through notifications or prompts.
  - Use cart abandonment reminders to re-engage users who haven’t completed their purchases.
  - Offer incentives for adding multiple items to the cart (e.g., free shipping thresholds).

---

### **2. Increase Session Duration**
- **Why?** Users with longer sessions are statistically more likely to convert.
- **Action Steps:**
  - Add personalized recommendations and curated product collections to extend browsing time.
  - Improve navigation and site speed to reduce friction during the shopping experience.
  - Introduce features like "Recently Viewed" or "You May Also Like" to encourage more engagement.

---

### **3. Boost Product Views**
- **Why?** Users who explore more products tend to have a higher likelihood of purchase.
- **Action Steps:**
  - Highlight popular products on the homepage to catch attention.
  - Use algorithms to suggest products similar to those already viewed.
  - Display cross-sell suggestions on product pages to promote complementary items.

---

### **4. Target Optimal Price Points**
- **Why?** Our analysis shows low to mid-range priced items (<$300) have the highest conversion likelihood.
- **Action Steps:**
  - Offer discounts or promotions on mid-range products to boost conversions.
  - Bundle products to create value-driven offers for users.
  - Use dynamic pricing to align prices with user demand and behavior.

---

### **5. Encourage Category Exploration**
- **Why?** Users engaging with diverse product categories are more likely to purchase.
- **Action Steps:**
  - Promote multi-category bundles (e.g., "Complete Your Look" or "Best Deals in Every Category").
  - Target users with ads or emails showcasing products from less explored categories.
  - Highlight diverse product ranges in campaigns to appeal to varied user interests.

---

### **Limitations**
- **Assumptions in Data Distribution:** The analysis assumes that the purchase decline is distributed randomly across sessions, categories, and price points. This may not capture systemic biases or specific external factors affecting user behavior (e.g., seasonal trends or market conditions).

- **Incomplete Data:** The dataset may include noise or missing values, particularly in categorical fields (e.g., brand, category_code). While cleaning and deduplication were performed, the impact of incomplete or inaccurate data on model performance cannot be ruled out.

- **Model Simplifications:** Predictive models were built using selected features, which may overlook additional factors (e.g., external economic indicators, promotional campaigns) that could influence cart-to-purchase behavior.

- **Simulation Limitations:** The simulated April dataset assumes the success of interventions (e.g., promoting high-likelihood sessions) without considering real-world constraints like user engagement limits or marketing resource availability.

## **Conclusion**
To address the decline in the cart-to-purchase ratio, the focus should be on recovering $213,921 in lost revenue through enhanced user engagement, optimized pricing, and category diversity. 

Immediate actions include cart abandonment campaigns, showcasing bestsellers, and cross-selling prompts, while medium-term strategies involve dynamic pricing, curated promotions, and category exploration. Collaboration across teams and investments in technology and marketing will ensure successful implementation.

These efforts aim to recover lost revenue, add ~$217,000 in monthly revenue, and foster long-term user engagement and growth.